Eliminating object prior-bias from sparse-projection tomographic reconstructions

### 2021-08-20 paper/post_tci.tex
changes in text in abstract, intro, related-work, Fig.1, Fig.2, and marked text changes in old TCI

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
  Tomographic reconstruction from undersampled measurements is a
  necessity when the measurement process is potentially harmful, needs
  to be rapid, or is resource-expensive. In such cases, information
  from previously existing longitudinal scans of the same object
  (`object-prior'), helps in the reconstruction from the current
  measurements of that object (`test'), while requiring significantly
  fewer updating measurements. In this work, we improve the state of
  the art by proposing the context under which priors can be
  effectively used based on the final goal of the application at hand.

  Our work is based on longitudinal data acquisition scenarios where
  we wish to study new changes that evolve within an object over time,
  such as in repeated scanning for disease monitoring, or in
  tomography-guided surgical procedures. While this is easily feasible
  when measurements are acquired from a large number of projection
  angles (`views'), it is challenging when the number of views is
  limited.

  If the goal is to track the changes while simultaneously reducing
  sub-sampling artefacts, we propose (1) acquiring measurements from
  an \emph{extremely small} number of views and using a
  \emph{`uniform'} prior-based reconstruction. If the goal is to
  observe \textbf{details} of new changes, we propose (2) acquiring
  measurements from a \emph{moderate} number of views (albeit, still
  sub-Nyquist), and using a more involved reconstruction routine. We
  show that in the latter case, a `spatially-varying' technique is
  appropriate in order to prevent the prior from adversely affecting the
  reconstruction of new structures that are absent in any of the
  earlier scans. The reconstruction of new regions is safeguarded from
  the bias of the prior by computing regional weights that moderate
  the local influence of the priors. We are thus able to effectively
  reconstruct both the old and the new structures in the test. We have
  tested the efficacy of our method on synthetic as well as real
  projection data. The results demonstrate the use of both uniform and
  spatially-varying priors in different scenarios. Our methods
  significantly improve the overall quality of the reconstructed data
  while minimizing the number of measurements needed for imaging in
  longitudinal studies.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Limited-view tomographic reconstruction, compressed sensing, priors, longitudinal studies.
\end{IEEEkeywords}

@@

In conventional data acquisition techniques, the tomographic
measurements $\boldsymbol{y}$ are acquired by sampling the physical
object $\boldsymbol{x}$ uniformly with a substantial number of views,
ideally above the Nyquist rate. In such a case when there are
sufficient measurements, reconstruction using the conventional
filtered backprojection (FBP) suffices, as seen in
Fig.~\ref{fig:story} (first column, 600 views). The figure shows the
ground truth image (260$\times$260) of naturally growing sprouts at the top
left (the details of this dataset is postponed to
Section~\ref{sec:results_uniform_prior}).

However, in the last decade, reconstruction from reduced views has
been made possible by assuming the data to exhibit sparsity under
certain mathematical transforms $\boldsymbol{\Upsilon}$ such as the
wavelet transforms, or the Discrete Cosine Transform (DCT). This is
known as sparsity prior and is the fundamental principle in the widely
used Compressed Sensing (CS) technique~\cite{Donoho,introCS}.  There
are multiple ways to incorporate the sparsity prior using CS. We use
the LASSO (least absolute shrinkage and selection operator) which
iteratively solves for the solution by penalizing a combination of
least squares error and $L_1$-norm of the sparse coefficients
$\boldsymbol{\theta}$ of the object $\boldsymbol{x}$. If $\boldsymbol{\Psi}$
represents the sparsity basis, i.e., if $\boldsymbol{x} = \boldsymbol{\Psi\theta}$, and 
$\boldsymbol{\mathcal{R}}$ represents the acquisition model, then the
LASSO solution is described by one that minimizes

 \begin{equation}
   TV(\boldsymbol{x}) = \sum_{i,j}\nabla_{ij}(\boldsymbol{x})
   \label{Eq:definition_TV}
\end{equation}

 \begin{equation}
   J_{TV}(\boldsymbol{x}) = \lVert\boldsymbol{\mathcal{R}x}- \boldsymbol{y}\rVert_2^2 + \lambda_{TV}TV(\boldsymbol{x})
   \label{Eq:simple_TV}
   \end{equation}
 Fig.~\ref{fig:story}
 (second column) demonstrates the benefit of using sparsity prior when
 the number of projection views is limited (100 views).
 We solve this cost function using the popular l1-regularized least
 squares (l1-ls) package available in~\cite{l1ls}, and choose DCT as
 our sparsity basis $\boldsymbol{\Upsilon}$. We had observed no significant improvement if, say, the wavelet transform were instead used as the basis.

@@

 When the number of views is significantly reduced (this paper), the
 sparsity prior alone is not sufficient.  In such cases additional
 information (`object-prior') specific to the current
 object being scanned (called the `test' henceforth) is useful in
 further improving the reconstruction.  While this has been done in
 the context of dynamic CT-scans (discussed more in
 Section~\ref{sec:related}), in this paper we use a set $S$ of
 previous scans of the same object. This is relevant in a longitudinal
 context: the acquisition of sequential CT scans of the same subject
 in order to track time-evolving changes within the subject's
 interior. As seen in Fig.~\ref{fig:story} (last column, 30 views) our
 method uses the set $S$ \emph{uniformly} by creating an
 eigenspace representation (Section~\ref{sec:method_uniform_prior}).

 However, with the use of such an object prior, a new challenge
 emerges. The prior set $S$ may potentially overwhelm the necessary
 details, and when several prior scans are available, finding the
 right prior is an issue. We therefore seek an algorithm that
 estimates the location and magnitude of new changes in the
 (unknown) test. As we show in
 Section~\ref{sec:method_spatially_varying_prior}, this eventually
 prevents the prior from adversely affecting the reconstruction of new
 regions in the test. We refer to this method as a
 \textit{spatially-varying} prior-based reconstruction routine, that
 still uses \textit{all} the previous scans in the set $S$, without
 using simply the prior scan, or endeavoring to find the ``right''
 prior.

 This paper is organized as follows. After discussing related work in
 Section~\ref{sec:related}, Section~\ref{sec:contributions} lays down
 the key contributions of this work. Details of the datasets created and
 used for validation
 are described in Section~\ref{sec:datasets}. Section~\ref{sec:tmh}
 demonstrates the utility of both the uniform and spatially-varying
 methods on a longitudinal medical dataset. We now move to the
 details. Section~\ref{sec:method_uniform_prior} describes the
 construction of the uniform eigenspace prior, followed by the
 corresponding results in real and synthetic 3D biological datasets in
 Section~\ref{sec:results_uniform_prior}. Section~\ref{sec:method_spatially_varying_prior}
 describes how the uniform prior needs to be modified when accurate
 details of new changes are to be observed. A spatially-varying
 technique offers a solution, the results of which are presented in
 Section~\ref{sec:results_spatially_varying_prior}. Section~\ref{sec:discussion}
 discusses tuning of the hyperparameters involved, and limitations of
 our method. Finally, we conclude with key inferences that can be
 drawn from our work in Section~\ref{sec:conclusions}.
% We present results on real and simulated 3D
%  data for different longitudinal studies.  These are studies wherein
%  sequential CT scans of the same object are acquired to track
%  time-evolving changes within its interior.

@@

 \section{Related Work}
 \label{sec:related}
 The idea of using a reduced number of views is most pronounced in
specialized applications. For example, in~\cite{Essam2015},
sparsity-constrained optimization is presented for angiography. Here,
the regions of interest are the vessels alone and they are highlighted
by physically inserting a contrast agent, and therefore there is an
inherent sharper contrast between the vessels and the background.  In
other applications where the spatial-gradient of the underlying volume
is known to be sparse, the `total variation' method, as used
in~\cite{Li2015,Polak2017}, can be applied.  In a more generalized
application scenario, like the ones addressed in this paper, any
information about the regions of interest is not known, and such
techniques have limited applicability.

@@

However, all these
techniques~\cite{daniil2015,koen2020,Van2015,HaoGao,Van2014}
essentially reconstruct all the different stages of the object
simultaneously, and use the extra redundancy across time
simultaneously. In the scenario of a longitudinal study (where
projection measurements are acquired at instants that are several
weeks apart) which we consider in this paper, such approaches are not
feasible. A longitudinal study will necessarily require reconstruction
results after acquisition of each set of measurements

There exists methods for estimating the new changes (in the test)
directly in the projection space. In such techniques which can
be used in longitudinal studies, the object prior
is projected in the same set of views as the current set of test
measurements, and a set of projection differences is computed. These
projection differences are essentially tomographic projections of the
difference between the unknown test and the known object prior. Hence,
a variety of reconstruction algorithms can be used to estimate such a
difference image which can then be added to the object prior to yield
the final estimate of the test. Such methods have been proposed
in~\cite{Pourmorteza2015,Lee2012}. However, the reconstruction of the
difference image will inherently contain sub-sampling artefacts, and
these artefacts will appear in the final reconstruction as there is no
mechanism to mitigate them (unlike the method that we propose). This
can be seen in Sec.~4 of our supplementary material~\cite{supp_paper},
where we have compared our method with~\cite{Lee2012}.

In this paper, we present a technique that makes use of multiple
object priors, corresponding to high quality reconstructions of
similar objects across time, typically in a longitudinal study.
We use these priors to improve the
reconstruction of the test from largely undersampled measurements. In
particular, we define the regions of change between the test and the
eigenspace spanned by the previous high quality object priors in a
purely data driven manner. Additionally, our technique differentiates
between genuine structural changes and ``changes'' that appear due to
undersampling artefacts (as seen in Sec.4 of our supplementary
material~\cite{supp_paper}).  Besides this, since we use a statistical
model with multiple object priors, we avoid the problem of selecting
an appropriate single object prior
unlike~\cite{PICCS,Pourmorteza2015,Lee2012,pirple,Marjolein2016} and
also make use of the additional information available in the multiple
priors. Our technique also has the advantage of not requiring any
expert-domain knowledge about the object being scanned.

@@

In our experiments, we generate parallel beam measurements from 2D
slices from each of the 8 volumes. Note that all these 8 slices are
located at the \emph{same} index (slice number corresponding to the
same depth) within each of the respective volumes.  Observe (in
Fig.~\ref{fig:RFA2_test_object-prior}) that the needle is seen in all
of the first 7 slices and the effect of ablation is seen in the 8th
slice.
%In our experiments using this dataset, we demonstrate the usefulness of \textit{both} uniform prior and spatially varying prior. For the first case, we choose the first 6 slices to build the object prior and use slice 7 as the test. For the latter case, we choose the first 7 slices to build the object prior and use slice 8 as the test.
%--------------------------Results on TMH data--------------------------
%------------------RFA 2 data
\begin{figure}[h!]
    \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template1.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 1}
    \end{subfigure}
    \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template2.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 2}
    \end{subfigure}
     \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template3.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 3}
    \end{subfigure}
       \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template4.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 4}
    \end{subfigure}
       \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template5.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 5}
    \end{subfigure}
              \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template6_color.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 6}
    \end{subfigure}
       \begin{subfigure}[b]{0.23\linewidth}
        \fcolorbox{white}{blue}{\includegraphics[width=\textwidth]{../images/tmh/RFA2/template7.png}}
%\captionsetup{labelformat=empty}       
 \caption{slice 7}
    \end{subfigure}     
     \caption{Radio-frequency ablation dataset: One of the slices (512$\times$512) from each of the 8 scan volumes of a longitudinal study dataset of the liver. Note that in volumes (a) through (g), the needle (shown in red in (f)) approaches the target tumor.} 
\label{fig:RFA2_test_object-prior}
\end{figure}

@@

%---------------------------------Uniform section---------------------
\section{Uniform prior-based reconstruction}
\label{sec:method_uniform_prior}
Having presented the application, we first review the algorithm~\cite{my_dicta_paper} for a uniform  prior-based reconstruction since we are extending this algorithm to 3D reconstructions as part of this work.
%Principal Component Analysis (PCA) has been traditionally used to find the significant modes of (Gaussian corrupted) data. In this regard, it has been widely applied in the context of data compression. However, PCA can also be seen as a tool to provide an orthogonal basis to represent the space in which most of the test data could lie (except the new changes). This space is constructed from the available set of previously scanned objects which must cover a realistically representative range of structures. %, which was shown~\cite{my_dicta_paper} to be better when compared to dictionary-based priors. %A preliminary version of this particular method was earlier presented in \cite{my_dicta_paper}. 
We first present the eigenspace-cum-CS prior-based reconstruction. To begin with, when an object is scanned multiple times, a set of high quality reconstructions (i.e., reconstructions from a dense set of projection views) may be chosen as object-prior for the reconstruction of future scan volumes, which in turn, may be scanned using far fewer measurements. The eigenspace $E_{\text{high}}$ of the $L$ previously scanned objects $Q_1,Q_2,...,Q_L$ is pre-computed. Here, it is assumed that most of the test volume, barring the new changes, can be expressed as a sparse linear combination of the principal components (eigenvectors of the covariance matrix) obtained from a group of structurally similar volumes. Hence, the object-prior is represented by means of PCA. For the eigenspace to encompass a range of possible structures in the test slice, the object-prior must represent a wide structural range. Moreover, if these volumes are not aligned, then they must be first registered before computing the prior. The prior is built by computing the covariance matrix from the template set $\{Q_i\}_{i=1}^L$. The space spanned by the eigenvectors $\{\boldsymbol{V_k}\}_{k=1}^{L-1}$ (eigenspace) of the covariance matrix is the object prior and is assumed to contain most of the test slice that is similar, but not necessarily identical to the object-prior. We use all of the $L-1$ orthogonal eigenvectors as a basis to represent the unknown test volume. Let $\boldsymbol{\mu}$ denote the mean of the previously scanned objects, and $\boldsymbol{\alpha}$ the unknown vector of eigen-coefficients of the test scan, of which $\alpha_k$ is the $k^{\textrm{th}}$ element. Then, once the eigenspace is pre-computed, the test is reconstructed by minimizing the following cost function:

@@

 \section{Results: Reconstruction by uniform prior}
 \label{sec:results_uniform_prior}

 The proposed method has been validated on new scans of biological
 specimens in a longitudinal setting. 
 %We emphasize that in
% most of the literature on tomographic reconstruction, the results are
 %shown on reconstruction from projections \emph{simulated} from 3D
 %volumes. This is because most CT scanners do not reveal the raw
 %projections, and instead output only the full reconstructed
 %volumes. Moreover, the process of conversion from the projections to
 %the full volumes is proprietary.

%In all figures in this section, `uniform prior' refers to optimizing Eq.~\ref{eq:spatially-varying_prior} with $\boldsymbol{W}(x,y,z)=1$.
%----------------------------------------------------------
%-------------------------------------------------
\subsubsection{\textbf{Okra}}
\label{Sec:okra_uniform}
The test volume was reconstructed from a partial set of
45 projections, i.e., $10\%$ of the projection views from which ground
truth was reconstructed.
 One of the slices of the reconstructed volumes is shown in
 Fig.~\ref{fig:okra_3D_results_uniform}. The SSIM values presented in
 Table~\ref{table:okra_ssim_full_volume} show that the presence of a
 uniform object prior significantly improves the overall
 reconstruction in comparison with the usage of sparsity prior alone.
 The selected 3D ground truth of template volumes, the test volume as
 well as the 3D reconstructions can be seen in the supplementary
 material~\cite{supp_paper}.

%----------------------------------------------------
\subsubsection{\textbf{Sprouts}}
\label{Sec:sprouts}
The test volume was reconstructed from partial set of 45
projections, i.e., $2.5\%$ of the projection views from which ground
truth was reconstructed.
One of the slices of the reconstructed
 volumes is shown in
 Fig.~\ref{fig:sprouts_3D_results_uniform}. Table~\ref{table:sprouts_ssim_full_volume}
 shows the SSIM of the full reconstructed volumes.  Here again, the
 presence of a uniform object prior significantly improves the overall
 reconstruction in comparison with the usage of sparsity prior alone.
 The selected 3D ground truth of template volumes, the test volume as
 well as the 3D reconstructions can be seen in the supplementary
 material~\cite{supp_paper}.

%----------------------------------------------------------------------
\section{Spatially-Varying prior-based reconstruction}
\label{sec:method_spatially_varying_prior}

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
  Tomographic reconstruction from undersampled measurements is a
  necessity when the measurement process is potentially harmful, needs
  to be rapid, or is resource-expensive. In such cases, information
  from previously existing longitudinal scans of the same object
  (`object-prior') helps in the reconstruction of the current object (`test')  from its
  significantly fewer measurements. A common problem with these techniques is
  the strong influence of object-priors in the reconstruction of new regions in the test.
  In this work, we mitigate this problem by first estimating the location of new regions and
  then imposing object-prior in \textit{only} the other old regions which are similar to the prior.

  Our work is based on longitudinal data acquisition scenarios where
  we wish to study new changes that evolve within an object over time,
  such as in repeated scanning for disease monitoring, or in
  tomography-guided surgical procedures. While this is easily feasible
  when measurements are acquired from a large number of projection
  angles (`views'), it is challenging when the number of views is
  limited (sub-Nyquist). We
  show that in the latter case, a `spatially-varying' technique is
  appropriate in order to prevent the prior from adversely affecting the
  reconstruction of new structures that are absent in any of the
  earlier scans. The reconstruction of new regions is safeguarded from
  the bias of the prior by computing regional weights that moderate
  the local influence of the priors. We are thus able to effectively
  reconstruct both the old and the new structures in the test. We have
  tested the efficacy of our method on synthetic as well as real
  projection data, in both 2D and 3D. Our method
  significantly improves the overall quality of reconstructed data
  while minimizing the number of measurements needed for imaging in
  longitudinal studies.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Limited-view tomographic reconstruction, regularization priors, object priors, longitudinal studies.
\end{IEEEkeywords}

@@

In conventional data acquisition techniques, tomographic
measurements $\boldsymbol{y}$ are acquired by sampling the physical
object $\boldsymbol{x}$ uniformly with a substantial number of views,
ideally above the Nyquist rate. In such a case when there are
sufficient measurements, reconstruction using conventional
filtered backprojection (FBP) suffices, as seen in
Fig.~\ref{fig:story} (first column, 600 views). The figure shows the
ground truth image (260$\times$260) of naturally growing sprouts at the top
left (the details of this dataset is postponed to
Section~\ref{sec:datasets}).

However, reconstruction from reduced views has
been made possible by assuming the data to exhibit certain properties such as
smoothness in image space, sparsity in gradient space~\cite{TV} or sparsity under
certain mathematical transforms~\cite{Donoho,introCS} such as the Discrete Cosine Transform (DCT).  There are multiple such regularization priors and our work aims at significantly improving
the reconstructions obtained from any of these methods. Hence, although our technique is agnostic to the type of regularization prior used, here we have used the total Variation (TV) method as part of our presented solution. If $\boldsymbol{\mathcal{R}}$ represents the acquisition model, then TV-regularized solution is described by one that minimizes a combination of
least squares error and $TV$-norm  of the object $\boldsymbol{x}$. This is described by

 \begin{equation}
   J_{TV}(\boldsymbol{x}) = \lVert\boldsymbol{\mathcal{R}x}- \boldsymbol{y}\rVert_2^2 + \lambda_{TV}TV(\boldsymbol{x})
   \label{Eq:simple_TV}
 \end{equation}

 where the $TV$ norm is given by

  \begin{equation}
   TV(\boldsymbol{x}) = \sum_{i,j}\nabla_{ij}(\boldsymbol{x}).
   \label{Eq:definition_TV}
\end{equation}

 Fig.~\ref{fig:story}
 (second column) demonstrates the benefit of using a regularization prior when
 the number of projection views is limited (100 views).
 We solve this cost function using an optimal first-order method for large scale TV regularization
 presented in~\cite{TVReg}  and available in~\cite{TVReg-lib}.

@@

 When the number of views is significantly reduced (this paper), the
 regularization prior alone is not sufficient.  In such cases additional
 information (`object-prior') specific to the current
 object being scanned (called the `test' henceforth) is useful in
 further improving the reconstruction.  While this has been done in
 the context of dynamic CT-scans (discussed more in
 Section~\ref{sec:related}), in this paper we use a set $S$ of
 previous scans of the same object. This is relevant in a longitudinal
 context: the acquisition of sequential CT scans of the same subject
 in order to track time-evolving changes within the subject's
 interior. As seen in Fig.~\ref{fig:story} (last column, 30 views) our
 method uses the set $S$ by creating an
 eigenspace representation (Section~\ref{sec:method_spatially_varying_prior}).

 However, with the use of such an object prior, a new challenge
 emerges. The prior set $S$ may potentially overwhelm the necessary
 details, and when several prior scans are available, finding the
 right prior is an issue. We therefore seek an algorithm that
 estimates the location and magnitude of new changes in the
 (unknown) test. As we show in
 Section~\ref{sec:method_spatially_varying_prior}, this eventually
 prevents the prior from adversely affecting the reconstruction of new
 regions in the test. We refer to this method as a
 \textit{spatially-varying} prior-based reconstruction routine, that
 still uses \textit{all} the previous scans in the set $S$, without
 using the prior scan \textit{as is}, or endeavoring to find the ``right''
 prior.

 This paper is organized as follows. After discussing related work in
 Section~\ref{sec:related}, Section~\ref{sec:contributions} lays down
 the key contributions of this work. Details of the datasets created and
 used for validation
 are described in Section~\ref{sec:datasets}. Section~\ref{sec:tmh}
 demonstrates the utility of our
 method on a longitudinal medical dataset. We now move to the
 details. Section~\ref{sec:method_spatially_varying_prior}
 describes the
 construction of a uniform eigenspace prior and computation of a spatially-varying map in order to preserve new changes in the test. 
 The results of are presented in
 Section~\ref{sec:results_spatially_varying_prior}. Section~\ref{sec:discussion}
 discusses tuning of the hyperparameters involved, and limitations of
 our method. Finally, we conclude with key inferences that can be
 drawn from our work in Section~\ref{sec:conclusions}.
% We present results on real and simulated 3D
%  data for different longitudinal studies.  These are studies wherein
%  sequential CT scans of the same object are acquired to track
%  time-evolving changes within its interior.

@@

 \section{Related Work}
 \label{sec:related}
 The idea of using a reduced number of views is most pronounced in
specialized applications. For example, in~\cite{Essam2015},
sparsity-constrained optimization is presented for angiography. Here,
the regions of interest are the vessels alone and they are highlighted
by physically inserting a contrast agent, and therefore there is an
inherent sharper contrast between the vessels and the background.  In
other applications where the spatial-gradient of the underlying volume
is known to be sparse, the `total variation' method, as used
in~\cite{Li2015,Polak2017}, can be applied.  However, as shown earlier
in Fig.~\ref{fig:story}, these regularization priors are not sufficient when very few
views are acquired, thus needing information from stronger priors such as those obtained from previous scans of the same or similar object.

@@

However, all these
techniques~\cite{daniil2015,koen2020,Van2015,HaoGao,Van2014}
essentially reconstruct all the different stages of the object
simultaneously, and use the extra redundancy across time
simultaneously. In the scenario of a longitudinal study (where
projection measurements are acquired at instants that are several
weeks apart) which we consider in this paper, such approaches are not
feasible. A longitudinal study will necessarily require reconstruction
results after acquisition of each set of measurements.

There exists methods for estimating the new changes (in the test)
directly in the projection space. In such techniques which can
be used in longitudinal studies, the object prior
is projected in the same set of views as the current set of test
measurements, and a set of projection differences is computed. These
projection differences are essentially tomographic projections of the
difference between the unknown test and the known object prior. Hence,
a variety of reconstruction algorithms can be used to estimate such a
difference image which can then be added to the object prior to yield
the final estimate of the test. Such methods have been proposed
in~\cite{Pourmorteza2015,Lee2012}. Here the reconstruction of the
difference image will inherently contain sub-sampling artefacts, and
these artefacts will appear in the final reconstruction as there is no
mechanism to mitigate them (unlike the method that we propose). This
can be seen in Sec.~4 of our supplementary material~\cite{supp_paper},
where we have compared our method with~\cite{Lee2012}.

In this paper, we present a technique that makes use of multiple
object priors, corresponding to high quality reconstructions of
similar objects across time, typically in a longitudinal study.
We use these priors to improve the
reconstruction of the test from largely undersampled measurements. In
particular, we define the regions of change between the test and the
eigenspace spanned by the previous high quality object priors in a
purely data driven manner. Additionally, our technique differentiates
between genuine structural changes and ``changes'' that appear due to
undersampling artefacts (as seen in Sec.4 of our supplementary
material~\cite{supp_paper}).  Besides this, since we use a statistical
model with multiple object priors, we avoid the problem of selecting
an appropriate single object prior
unlike~\cite{PICCS,Pourmorteza2015,Lee2012,pirple,Marjolein2016} and
also make use of the additional information available in multiple
priors. Our technique also has the advantage of not requiring any
expert-domain knowledge about the object being scanned.

@@

In our experiments, we generate parallel beam measurements from 2D
slices from each of the 8 volumes. Note that all these 8 slices are
located at the \emph{same} index (slice number corresponding to the
same depth) within each of the respective volumes.  Observe (in
Fig.~\ref{fig:RFA2_test_object-prior}) that the needle is seen in all
of the first 7 slices and the effect of ablation is seen in the 8th
slice.
%In our experiments using this dataset, we demonstrate the usefulness of \textit{both} uniform prior and spatially varying prior. For the first case, we choose the first 6 slices to build the object prior and use slice 7 as the test. For the latter case, we choose the first 7 slices to build the object prior and use slice 8 as the test.
%--------------------------Results on TMH data--------------------------
%------------------RFA 2 data
\begin{figure}[h!]
    \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template1.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 1}
    \end{subfigure}
    \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template2.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 2}
    \end{subfigure}
     \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template3.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 3}
    \end{subfigure}
       \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template4.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 4}
    \end{subfigure}
       \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template5.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 5}
    \end{subfigure}
              \begin{subfigure}[b]{0.24\linewidth}
        \includegraphics[width=\textwidth]{../images/tmh/RFA2/template6_color.png}
%\captionsetup{labelformat=empty}       
 \caption{slice 6}
    \end{subfigure}
       \begin{subfigure}[b]{0.23\linewidth}
        \fcolorbox{white}{blue}{\includegraphics[width=\textwidth]{../images/tmh/RFA2/template7.png}}
%\captionsetup{labelformat=empty}       
 \caption{slice 7}
    \end{subfigure}     
     \caption{Radio-frequency ablation dataset: One of the slices (512$\times$512) from each of the 7 scan volumes of a longitudinal study dataset of the liver. Note that in volumes (a) through (g), the needle (shown in red in (f)) approaches the target tumor.} 
\label{fig:RFA2_test_object-prior}
\end{figure}

@@

%---------------------------------Uniform section---------------------
\section{Spatially-Varying prior-based reconstruction}
\label{sec:method_spatially_varying_prior}
Having presented the application, we first review the algorithm~\cite{my_dicta_paper} for a uniform  prior-based reconstruction since we are extending this algorithm to 3D reconstructions as part of this work.
%Principal Component Analysis (PCA) has been traditionally used to find the significant modes of (Gaussian corrupted) data. In this regard, it has been widely applied in the context of data compression. However, PCA can also be seen as a tool to provide an orthogonal basis to represent the space in which most of the test data could lie (except the new changes). This space is constructed from the available set of previously scanned objects which must cover a realistically representative range of structures. %, which was shown~\cite{my_dicta_paper} to be better when compared to dictionary-based priors. %A preliminary version of this particular method was earlier presented in \cite{my_dicta_paper}. 
We first present the eigenspace-cum-CS prior-based reconstruction. To begin with, when an object is scanned multiple times, a set of high quality reconstructions (i.e., reconstructions from a dense set of projection views) may be chosen as object-prior for the reconstruction of future scan volumes, which in turn, may be scanned using far fewer measurements. The eigenspace $E_{\text{high}}$ of the $L$ previously scanned objects $Q_1,Q_2,...,Q_L$ is pre-computed. Here, it is assumed that most of the test volume, barring the new changes, can be expressed as a sparse linear combination of the principal components (eigenvectors of the covariance matrix) obtained from a group of structurally similar volumes. Hence, the object-prior is represented by means of PCA. For the eigenspace to encompass a range of possible structures in the test slice, the object-prior must represent a wide structural range. Moreover, if these volumes are not aligned, then they must be first registered before computing the prior. The prior is built by computing the covariance matrix from the template set $\{Q_i\}_{i=1}^L$. The space spanned by the eigenvectors $\{\boldsymbol{V_k}\}_{k=1}^{L-1}$ (eigenspace) of the covariance matrix is the object prior and is assumed to contain most of the test slice that is similar, but not necessarily identical to the object-prior. We use all of the $L-1$ orthogonal eigenvectors as a basis to represent the unknown test volume. Let $\boldsymbol{\mu}$ denote the mean of the previously scanned objects, and $\boldsymbol{\alpha}$ the unknown vector of eigen-coefficients of the test scan, of which $\alpha_k$ is the $k^{\textrm{th}}$ element. Then, once the eigenspace is pre-computed, the test is reconstructed by minimizing the following cost function:

@@

%----------------------------------------------------------------------
